---
title: "Mangrove monitoring"
author: "Morris Ngowa"
date: "2025-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Load Necessary Libraries
```{r}
library(xlsx)
library(dplyr)
library(ggplot2)
library(tidyr)
library(vegan)
library(BIOMASS)
```


## 2. Load Dataset
```{r}
df <- read.xlsx("mangrove_seedling.xlsx", sheetName = "raw_data", colIndex = 1:8)
head(df)
```

```{r}
## create factors
#df$Plot.Size <- as.factor(df$Plot.Size)
df$Site <- as.factor(df$Site)
df$Species <- as.factor(df$Species)
df$Plot.No <- as.factor(df$Plot.No)
head(df)
```

```{r}
# convert mm to cm for DBH whereplot size is 12.6, 1, 6.25
df <- df %>% mutate(DBHcm = ifelse(Plot.Size == 100, DBH, DBH/10))
head(df)
```
```{r}
# export data to csv 
# write.csv(df, file = "mangrove_data.csv", row.names = FALSE)
```


## 3. Species Distribution Across Sites
```{r}
species_distribution <- df %>% 
  group_by(Site, Species) %>% 
  summarise(Total_Count = sum(Count, na.rm = TRUE) ) %>% 
  arrange(Site, desc(Total_Count))

print(species_distribution)
```

Groups data by Site and Species.
Computes the total count of each species at different sites.
Arranges species in descending order per site.


## 4. Tree Density Per Species Per Site
```{r}
density_per_site <- df %>% 
  group_by(Site, Species) %>% 
  summarise(Density = sum(Count) / sum(Plot.Size), .groups = 'drop')

print(density_per_site)
```

Calculates tree density (total count divided by total plot size).

## 5. Summary Statistics for DBH (Diameter at Breast Height) and Height

```{r}
summary_stats <- df %>% 
  group_by(Species) %>% 
  summarise(
    Mean_DBH = mean(DBHcm, na.rm = TRUE),
    Median_DBH = median(DBHcm, na.rm = TRUE),
    Mean_Height = mean(H, na.rm = TRUE),
    Median_Height = median(H, na.rm = TRUE),
    Count = n()
  )

print(summary_stats)
```

Computes mean and median DBH and height for each species.

## 6. Species Richness (Diversity per Plot)

```{r}
species_richness <- df %>% 
  group_by(Site, Plot.No) %>% 
  summarise(Richness = n_distinct(Species))

print(species_richness)
```

Measures species richness (the number of unique species in each plot).

## 7. Survival Rate Estimation
```{r}
survival_rate <- df %>% 
  group_by(Species) %>% 
  summarise(Survival_Rate = sum(Count, na.rm = TRUE) / n())

print(survival_rate)
```

Estimates survival rate by dividing the total count by the number of observations.

## 8. Growth Rate Estimation
```{r}
growth_rate <- df %>% 
  group_by(Species) %>% 
  summarise(Growth_Rate = mean(H / DBH, na.rm = TRUE))

print(growth_rate)
```

Growth rate is estimated using Height-to-DBH ratio.

## 9. Regeneration Rate Estimation
```{r}
regeneration_rate <- df %>% 
  group_by(Species, Site) %>% 
  summarise(Regeneration_Rate = sum(Count) / length(unique(Plot.No)), .groups = 'drop')

print(regeneration_rate)
```

Regeneration rate is calculated as total count per species divided by the number of unique plots.

## 10. Forest Density Calculation
```{r}
forest_density <- df %>% 
  group_by(Site) %>% 
  summarise(Forest_Density_stems_per_ha = sum(Count) / sum(Plot.Size)*10000)

print(forest_density)
```

Forest density is calculated as total tree count per site divided by total plot size.

## 11. Mangrove Biomass Estimation
```{r}
biomass <- df %>% 
  mutate(Biomass = 0.251 * (DBH^2.46)) %>% 
  group_by(Species, Site) %>% 
  summarise(Total_Biomass = sum(Biomass, na.rm = TRUE), .groups = 'drop')

print(biomass)
```


Uses an allometric equation to estimate biomass based on DBH.
Groups results by Species and Site.


## 12. Visualization - Species Distribution Across Sites
```{r}
ggplot(species_distribution, aes(x = reorder(Species, -Total_Count), y = Total_Count, fill = Site)) + 
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Mangrove Species Distribution across Sites", x = "Species", y = "Total Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Creates a bar plot to visualize the distribution of species across sites.
Key Insights from the Code
Species Richness & Distribution – Understand which species are dominant in different locations.
Survival Rate – Assess how well mangrove species are surviving.
Growth Rate – Identify fast-growing vs. slow-growing species.
Regeneration Rate – Evaluate the natural regeneration of mangrove forests.
Forest Density – Understand tree density across different sites.
Mangrove Biomass – Estimate total biomass, which is critical for carbon sequestration studies.
Would you like any modifications or additional analyses? 🚀